---
title: 'FunResp: Multicategory logit models for habitat selection including functional
  response'
author: "Claudia Dupke"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Vignette for the FunResp package}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
abstract: "Test Discrete choice models are implemented as mixed-effects
  baseline-category logit models to analyze the variation  of habitat category use 
  of moving animal over various temporal scales. Specifically, changes
  in habitat use with respect to habitat availability can be investigated leading to  
  functional response curves that further can be fitted to Holling's types. \n \n \n"
editor_options: 
  
  chunk_output_type: console
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=8.5, fig.height=5.5, fig.align='center', warning=FALSE, message=FALSE, cache = T)
```

## Getting FunResp
 
 from Github
 
```{r, eval = F}
 install.packages("devtools")
 library(devtools)
 install_github("cdupstats/FunResp")
```

 
##  Getting started
 
 Load packages and data
```{r, echo = T, message = F}
   library(FunResp)
   library(mgcv)
   data(Data)
   data(Landscapegrid)
```  
 
 Explore the data (for your own data use other packages for an in-depth analysis). 
 
```{r, echo = T, fig.width=8, fig.height=8}
   image(Landscapegrid,attr="habitats")
   points(Data$x,Data$y,col=as.numeric(Data$burst)+10)

   Data$month <- as.numeric(format(Data$date,"%m"))
   Data$hour <- as.numeric(format(Data$date,"%H"))
   table(Data[,c("month","burst")])
``` 
 
 Get landscape information
 
```{r}
  Data$habitat  <- over(SpatialPoints(cbind(Data$x,Data$y)),Landscapegrid)$habitat
  Data$habitat  <- relevel(Data$habitat, ref="old mixed")
``` 
## Setting up Multicategory logit models

Get availability of habitat types in monthly home ranges 
and merge with landscape data.

```{r, echo = T, fig.width=8, fig.height=8}
   par(mfrow=c(3,4),mar=c(1,1,1,1),oma=c(1,1,1,1))
   HRavail <- HRti(Data,varti="month",Landscapegrid,plot=TRUE)

   Data$month_f <- as.character(floor(Data$month))
   Data2 <- cbind(Data,HRavail[apply(Data[,c("burst","month_f")],1,paste,collapse="."),-c(1:2)])
   Data2$avail <- unlist(sapply(1:nrow(Data2),function(i)   HRavail[apply(Data2[i,c("burst","month_f")],1,paste,collapse="."),as.character(unlist(Data2[i,"habitat"]))]))
   Data2$relavail <- Data2$avail/Data2$total
```
  
   Prepare the model, by adding a dummy variable for the random effect and set the model formula.
   Basically, you can choose any model formula for each habitat type (see examples behind the #).
   
```{r}
   Data2$dum1 <- 1 #dummy variable
   # Run models (for all habitats but the baseline category)
   # set the model first
    f1 <- formula(present~offset(log(rel.avail))
                           +s(hour,bs="cp") #+s(month,bs="cp")#+te(month,hour,by=sex,bs="cp")
                           + rel.avail # s(rel.avail,bs="ps")#+rel.avail*season
                           +s(burst,by=dum1,bs="re"))
```

Check whether you have enough data for each habitat type. Here, we use a very small data set and I remove all habitats with less than 11 recordings.
 Then relevel the factor variable of the habitat types to the most common reference type, here: "old mixed"
 
```{r}
  nn <- table(Data2$habitat)
  habitats <- names(nn)[which(nn>10)] #take only those with more than 10 recordings
  Data2 <- subset(Data2,habitat %in% habitats)
  Data2$habitat  <- factor(relevel(Data2$habitat, ref="old mixed")) #make sure that baseline habitat is at first position like relevel(Data$habitat, ref="old mixed")
```
  
 Estimate the multicategory logit models, one for each habitat except the baseline category and plot the results. This is just a small toy data set for the purpose of demonstration. Therefore, the patterns here are somewhat disappointing. 
   
```{r, echo = T, fig.width=8, fig.height=8}
   MM <- multicat1(Data2,HRavail,f1,sp=c(1,2,1))
   names(MM) <- habitats[-1]
   par(mfrow=c(3,6),mar=c(1,1,1,1),oma=c(1,1,1,1))
   invisible(lapply(MM,plot))
```
 

## Functional response

Now you can use the model to make predictions for functional repsonse for each habitat type.
First, set the data for predictions as a sequence of values for relative availability for each habitat (make sure that it does not exceed the range of the observed data as we should avoid extrapolation).

```{r}
seqRA <- sapply(habitats[-1],function(ty) try(seq(min(exp(MM[[ty]]$model$"offset(log(rel.avail))")),min(quantile(exp(MM[[ty]]$model$"offset(log(rel.avail))"),.95),0.9), length=100)))
seqRA <- cbind(seqRA,"old mixed"=seq(0.001,0.5,length=nrow(seqRA)))
```

You also need to fix the values for availability for all available habitats as these are needed for predicting the use of a focal habitat. As you remember, we assume that the animal can choose between different habitats at the same time (discrete choice) and therefore the probability of use of other habitats affect the probability of use of the focal habitat. As we told our model, the probability of use depends on the availability of a habitat. I recommend using the mean and transform the values so that availabilities of all habitats sum to 1. 

```{r}
#mean availability for each habitat
UP <- unique(Data2[,c("burst","habitat","relavail")])
ma3 <- aggregate(UP$relavail,by=list(UP$habitat),mean)
meanavail <- ma3$x
names(meanavail)  <- ma3$Group.1
meanavail <- meanavail/sum(meanavail)
```

As your model varies over time you have to fix the values for time and all other covariates included in the model to make the prediction.

```{r}
hour <- 0
#start predictions
Preds <- lapply(habitats[-1],makpred_P,hoset=hour,seqRA=seqRA,meanavail=meanavail,Mod=MM)
names(Preds) <- habitats[-1]
#plot predictions
par(mfrow=c(2,3),mar=c(4,4,1,1),oma=c(2,2,2,2))
invisible(lapply(1:length(Preds),function(X) plot(Preds[[X]][,"relavail"],Preds[[X]][,"fit"],main=names(Preds)[X],type="l",ylab="Proportion of use",xlab="Relative availability")))
```

### Holling's type

The estimated functional response curve are now linked to Holling's types: 
For type I:
$h_I(x)=a x$, where $x$ is the relative availability of a habitat, a value
between 0 and 1, and $a$ the proportionality factor; for type II:
 $h_{II} (x)=\frac{ax}{b+x}$ and for type III: $h_{III}
 (x)=\frac{ax^2}{b^2+x^2}$. The optimisation routine minimises the residual sum
 of squares between the values of $h_I(x)$, $h_{II}(x)$ and $h_{III}(x)$ and the 
 values of the curve estimated from the
 multicategory logit models by finding optimal values for $a$ and $b$. The R function
 \emph{optim} of the R-package \emph{stats} was used for this
 purpose, with the values of $a$ and $b$ limited to 0
 and 1 for $h_{II}$ and $h_{III}$.
 The Holling type with the smallest residual sum of squares is considered to
 best explain the shape of the functional response curve.
 
 ```{r, echo = T, fig.width=8, fig.height=4}
HT <- lapply(1:length(Preds), function(X) optimHolling(Preds[[X]][,"relavail"],Preds[[X]][,"fit"]))
names(HT) <- names(Preds)
HTty <- do.call(rbind, HT)

par(mfrow=c(2,3),mar=c(4,4,1,1),oma=c(2,2,2,2))
invisible(lapply(1:length(Preds),function(ty) {
			x <- Preds[[ty]][,"relavail"]	
			plot(x,Preds[[ty]][,"fit"],main=names(Preds)[ty],type="l",ylab="Proportion of use",xlab="Relative availability")
			if(HTty[ty,"type"]==1){
				lines(x,Holling1(x,HTty[ty,"a"]),col="red", lty=3, lwd=3)
			}
			if(HTty[ty,"type"]==2){
				lines(x,Holling2(x,HTty[ty,"a"],HTty[ty,"b"]),col="red",lty=3,lwd=3)
			}
			if(HTty[ty,"type"]==3){
				lines(x,Holling3(x,HTty[ty,"a"],HTty[ty,"b"]),col="red",lty=3,lwd=3)
			}
			abline(a=0,b=1,col="darkgrey")
			mtext(paste("Hollingtype",c("I","II","III","IV")[HTty[ty,"type"]],"\n a=",round(HTty[ty,"a"],2),"\n b=",round(HTty[ty,"b"],2)),side=1,line=-2,adj=0.9, col = "red")
		}))
  
```

One more example, based on real data:
 ```{r,  fig.width=8, fig.height=8}
data(FemalesNight)
par(mfrow=c(4,3),mar=c(4,4,1,1),oma=c(2,2,2,2))
invisible(lapply(1:length(FemalesNight),function(X) plot(FemalesNight[[X]][,"RA"],FemalesNight[[X]][,"fit"],main=names(FemalesNight)[X],type="l",ylab="Proportion of use",xlab="Relative availability")) )

HT <- lapply(1:length(FemalesNight), function(X) optimHolling(FemalesNight[[X]][,"RA"], FemalesNight[[X]][,"fit"]))
names(HT) <- names(FemalesNight)
HTty <- do.call(rbind, HT)

par(mfrow=c(4,3),mar=c(4,4,1,1),oma=c(2,2,2,2))
invisible(lapply(1:length(FemalesNight),function(ty) {
			x <- FemalesNight[[ty]][,"RA"]	
			plot(x,FemalesNight[[ty]][,"fit"],main=names(FemalesNight)[ty],type="l",ylab="Proportion of use",xlab="Relative availability")
			if(HTty[ty,"type"]==1){
				lines(x,Holling1(x,HTty[ty,"a"]),col="red", lty=3, lwd=3)
			}
			if(HTty[ty,"type"]==2){
				lines(x,Holling2(x,HTty[ty,"a"],HTty[ty,"b"]),col="red",lty=3,lwd=3)
			}
			if(HTty[ty,"type"]==3){
				lines(x,Holling3(x,HTty[ty,"a"],HTty[ty,"b"]),col="red",lty=3,lwd=3)
			}
			abline(a=0,b=1,col="darkgrey")
			mtext(paste("Hollingtype",c("I","II","III","IV")[HTty[ty,"type"]],"\n a=",round(HTty[ty,"a"],2),"\n b=",round(HTty[ty,"b"],2)),side=1,line=-2,adj=0.9, col = "red")
		}))
```


## Prediction of habitat use over temporal scales

 Now, you can use the model to predict habitat use over the range of covariates used in your model, e.g. time of day (hour). Keep in mind, that this is just a small toy data set for the purpose of demonstration. Therefore, the patterns here are somewhat disappointing. 
The use of the reference (or baseline) type has to be calculated seperately. 
 
```{r, echo = T, fig.width=8, fig.height=8}
hour <- 0:23
RA <- matrix(rep(meanavail,each=length(hour)),nrow=length(hour),dimnames=list(c(),c(names(meanavail))))
Preds <- lapply(habitats[-1],makpred_P,hoset=hour,seqRA=RA,meanavail=meanavail,Mod=MM)
names(Preds) <- habitats[-1]
par(mfrow=c(2,4))
invisible(lapply(1:length(Preds),function(X) plot(Preds[[X]][,"hour"],Preds[[X]][,"fit"],type="l",ylim=c(0,.4),main=names(Preds)[X],ylab="Proportion of use",xlab="Time of day (Hour)")))

# prediction for baseline category
predhour <- do.call(rbind,lapply(Preds,function(X) X[,"fit"]))
pred_base <- 1-apply(predhour,2,sum)

plot(hour,pred_base,type="l",ylim=c(0,.4),main=habitats[1],ylab="Proportion of use",xlab="Time of day (Hour)")
```


